Supplementary Material for ``Evaluating infectious disease forecasts with allocation scoring rules''

### 2023-12-14 alloscore_manuscript/alloscore-application.Rnw
fixes re issues #68, #80, #81, cite for regret, and a few sec:numerics edits

To make this a more easily interpreted measure of forecast performance, we will adjust the raw score by subtracting the minimum loss $l_{\text{oracle}}(y;K)$ achievable by an \emph{oracle} allocator which has precise foreknowledge of the outcomes $y_i$.
This minimum loss is given by $l_{\text{oracle}}(y;K) = L \cdot \max(0, \sum_{i=1}^{N}y_i - K)$: if there are enough resources to meet the total resource need ($\sum_{i=1}^{N}y_i \leq K$), the oracle can achieve a loss of 0, but if the resource constraint is less than the total needed resources ($\sum_{i=1}^{N}y_i > K$), it is not possible to do better than a loss of $L \cdot (\sum_{i=1}^{N}y_i - K)$.

Our scoring rule is then given by
\begin{align}
S_A(F, y; K) &= S_A^{\text{raw}}(F, y; K) - l_{\text{oracle}}(y;K) \\
& =
\begin{cases}
\sum_{i=1}^n L  \cdot  \max(0, y_i - x_i^{F,K}), &  \sum_{i=1}^{N}y_i \leq K \\
L \cdot K - L \cdot \sum_{i=1}^{N} \min(x_i^{F,K}, y_i), &  \sum_{i=1}^{N}y_i > K.
\end{cases}
\end{align}
In the first case, the oracle incurs no loss so that the raw and adjusted scores coincide. The second case can be read as taking the $K$ resource units perfectly allocated by the oracle as a base penalty on
the imperfect forecast $F$ and then, for each location $i$, reducing this penalty by however much of the need $y_i$
is met with the Bayes act component $x_i^{F,K}$.
The oracle adjustment aligns with the idea that \emph{opportunity loss} (often known as \emph{regret} or (negative) \emph{relative utility}) is often a more important quantity than absolute loss \elr{needs citation}.

\subsubsection{Integrating the allocation score across resource constraint levels}
\label{sec:methods.detailed.integrated_allocation}

@@

We have described in previous sections how both the allocation score and the weighted interval score (WIS) arise from a single standard procedure for developing proper scoring rules.
The allocation score arises when the decision relates to how a fixed quantity of resources should be allocated to multiple locations.
The WIS and CRPS arise when the decision relates to how much of a resource to order in the context of the cost of the resource and wanting to minimize unmet need in one or more locations.
In both settings, the optimal solution sets the resource level in each location to a quantile of the respective forecast distributions.

These decision making problems differ according to the challenge faced by the decision maker: a fixed constraint on the available resources for the allocation problem, or a cost per unit of resources in the resource purchasing problem. In fact, it is possible to combine these into a more general problem where the decision maker must decide on both the total level of resources to purchase and an allocation of those resources across multiple locations, subject to a cost per unit of resources and a constraint on the total quantity of resources that can be purchased. The quantile loss and the allocation score presented in this work are both special cases of the score that emerges from that more general decision making problem that integrates across resource cost and constraint parameters together. We pursue this direction further in other work that is in progress.

>>>>

To make this a more easily interpreted measure of forecast performance, we will adjust the raw score by subtracting the minimum loss $l_{\text{oracle}}(y;K)$ achievable by an \emph{oracle} allocator which has precise foreknowledge of the outcomes $y_i$.
When the oracle has sufficient resources to meet the total need, i.e., when $\sum_{i=1}^{N}y_i \leq K$, the oracle's loss
$L \cdot \max(0, \sum_{i=1}^{N}y_i - K) = 0$ and there is no adjustment.  Thus the allocation score coincides with
the raw score,
\begin{align}
S_A(F, y; K) = S_A^{\text{raw}}(F, y; K)=\sum_{i=1}^N L \cdot \max(0, y_i - x_i^{F,K}) \quad \text{if } \sum_{i=1}^{N}y_i \leq K.
\end{align}
On the other hand, when the oracle cannot cover all need and incurs a loss of $L \cdot \sum_{i=1}^{N}y_i - L\cdot K > 0$, we adjust
the raw score by this loss to get
\begin{align}
S_A(F, y; K)  &= S_A^{\text{raw}}(F, y; K) - L \cdot \sum_{i=1}^{N}y_i + L\cdot K  \nonumber\\
&= L \cdot \sum_{i=1}^{N} \left\{\max(0, y_i - x_i^{F,K}) - y_i\right\}  + L\cdot K \nonumber \\
&= L \cdot K - L \cdot \sum_{i=1}^{N} \min(x_i^{F,K}, y_i) \quad \text{if } \sum_{i=1}^{N}y_i > K.
\end{align}
This can be read as taking the $K$ resource units perfectly allocated by the oracle as a base penalty on
the imperfect forecast $F$ and then, for each location $i$, reducing this penalty by however much of the need $y_i$
is met with the Bayes act component $x_i^{F,K}$.
The oracle adjustment aligns with a common theme in economic decision theory that \emph{opportunity loss} (often known as \emph{regret} or (negative) \emph{relative utility}) is often a more important quantity than absolute loss (see e.g., \cite{DIECIDUE201788}).

\subsubsection{Integrating the allocation score across resource constraint levels}{}
\label{sec:methods.detailed.integrated_allocation}

@@

We have described in previous sections how both the allocation score and the weighted interval score (WIS) arise from a single standard procedure for developing proper scoring rules.
The allocation score arises when the decision relates to how a fixed quantity of resources should be allocated to multiple locations.
The WIS and CRPS arise when the decision relates to how much of a resource to order in the context of the cost of the resource and wanting to minimize unmet need in one or more locations.
In both settings, the score can be defined in terms of the shortage or excess of resource procurement levels against realized need where
these levels are expressable as quantiles of the respective forecast distributions by virtue of minimizing expected shortages or excesses.

These decision making problems differ according to the challenge faced by the decision maker: a fixed constraint on the available resources for the allocation problem, or a cost per unit of resources in the resource purchasing problem. In fact, it is possible to combine these into a more general problem where the decision maker must decide on both the total level of resources to purchase and an allocation of those resources across multiple locations, subject to a cost per unit of resources and a constraint on the total quantity of resources that can be purchased. When 
constraints are likely to be large, and therefore of less importance, a composite score for this general scenario would converge to the WIS or CRPS.  And if the constraints are likely to be tight, and therefore to be a determining factor, this composite score would converge to the IAS.  
We pursue this direction further in other work that is in progress.

### 2023-12-14 alloscore_manuscript/supplement.rnw
fixes re issues #68, #80, #81, cite for regret, and a few sec:numerics edits

	\item Descriptions of 
		\begin{todolist}		
			\item CRPS as average quantile score across $C \in [0/L]$ decision problems
			\item IS as average of two quantile scores with a prob-width penalty
			\item WIS as average quantile score across 23 C/L problems.
		\end{todolist}
	%item Sketch of scoring for decision problems involving both cost and constraint.
	\item  Derivation of case 2 in formula for Oracle adjustment
\end{todolist}

@@

Let $Y$ be a random variable
representing the future level of an undesirable outcome such as severe COVID incidence. Let $x \in \mathbb{R}_+$ be a decision variable
representing levels of some costly counter-measure, such as procurement of monoclonal antibody treatments,
that can be taken at a cost $C>0$ per unit in preparation for $Y$.\footnote{
Quantiles could also be derived for a problem in which $x$ and $Y$ take negative values, corresponding, for instance, to a decision
maker that both buys and sells in a resource market and a $Y$ that takes negative values when ``recoveries'' outnumber incidence.  
But we do not consider such scenarios in this work.	 
}  A decision maker must decide on a level $x$ of investment in the
counter-measure, and wishes to avoid excesses in either the expediture $Cx$ or the shortage $(y-x)_{+}$ when $Y=y$ is
realized. To formalize the trade-off between these potential excesses we quantify the loss associated with a unit of
shortage by a constant $L>C$ (which assumes that the counter-measure has some practical value) and combine the total shortage loss with expenditure into a \emph{loss function}\footnote
{This does involve a confusing use of the word \emph{loss} to refer to two different quantities, but this seems to be
an ingrained and unavoidable habit in the literature.}
\[
l(x,y) = Cx + L(y-x)_{+}.
\] 
The decision problem is then to select a random future loss $l(x,Y)$ in a way that aligns with the preference that $l(x,y)$ be
as low as possible given any realization $Y=y$.

To give the decision problem more structure we assume the decision maker either knows the distribution $F$ of $Y$, or
wishes to proceed as if a forecast $F$ of $Y$ were true. This gives us what is known in decision theory as a decision
problem \emph{under risk} (regarding the future value of $Y$) as opposed to one \emph{under uncertainty} where both $Y$
as well as $F$ are unknown when the decision is to be made. A principle commonly invoked in this situation\footnote
{Note that this priciple might be inappropriate when the decision maker is \emph{risk averse} in some way such as
having a preference for random losses with lower variance.} is that the decision maker should or will seek to minimize the expected
loss
\begin{align}
  \mathbb{E}_{F}[l(x,Y)] = Cx + L\mathbb{E}_{F}[(Y-x)_{+}]. \label{eqn:expQloss}
\end{align}

@@

\section{Numerical computation of allocation scores}
\label{sec:numeric}

Exact computation of the allocation score $S_A(F, y; K)$ of a forecast $F$ requires an explicit choice of and expression for a 
solution $x^{F,K} = \bm{q}_{\tau^{\star},F} \in \undertilde{\bm{F}}^{-1}(\tau^{\star})$ to \eqref{eqn:quantiles-sum-to-K} where
$\tau^{\star}$ solves \eqref{eqn:tau-inclusion}. For practical scenarios in which the forecast $F$ has a form preventing 
analytical solutions to \eqref{eqn:quantiles-sum-to-K}, we seek a method of generating an approximation $\tilde{\tau}$ to 
a solution $\tau^{\star}$ which offers control (at least in principle) over a robust measure of scoring rule error such as
the sum of absolute coordinate errors
\begin{align}
\mathrm{SAE}(Y,\tilde{\tau}) = \sum_{i=1}^{N} \left| (Y_i-q_{\tilde{\tau},F_i})_+ - (Y_i-x_i^{F,K})_+ \right|
\end{align}
% of the Bayes risk error or ``optimality gap''
% \begin{align}
% \left|\mathbb{E}_F\left[s_A(\bm{q}_{\tilde{\tau},F}, Y)\right] - \mathbb{E}_F[s_A(x^{F,K}, Y)]\right| = 
% L\left| \sum_{i=1}^{N}\mathbb{E}_{F_i}\left[(Y_i-q_{\tilde{\tau},F_i})_+ - (Y_i-x_i^{F,K})_+\right]\right|
% \end{align}
where $\bm{q}_{\tilde{\tau},F}$ solves the linear program
$\sum_{x \in \undertilde{\bm{F}}^{-1}(\tilde{\tau})}x_i = K$.

>>>>

	\item Descriptions of 
		\begin{todolist}		
			\item CRPS as average quantile score across $C \in [0/L]$ decision problems
			\item IS as average of two quantile scores with a prob-width penalty
			\item WIS as average quantile score across 23 C/L problems.
		\end{todolist}
	%item Sketch of scoring for decision problems involving both cost and constraint.
	\item[\xmark]  Derivation of case 2 in formula for Oracle adjustment (done in manuscript?)
\end{todolist}

@@

Let $Y$ be a random variable
representing the future level of an undesirable outcome such as severe COVID incidence. Let $x \in \mathbb{R}_+$ be a decision variable
representing levels of some costly counter-measure, such as procurement of monoclonal antibody treatments,
that can be taken at a cost $C>0$ per unit in preparation for $Y$.\footnote{
Quantiles could also be derived for a problem in which $x$ and $Y$ take negative values, corresponding, for instance, to a decision
maker that both buys and sells in a resource market and a $Y$ that takes negative values when ``recoveries'' outnumber incidence.  
But we do not consider such scenarios in this work.	 
}  A decision maker must decide on a level $x$ of investment in the
counter-measure, and wishes to avoid excesses in either the expediture $Cx$ or the shortage $(y-x)_{+}$ when $Y=y$ is
realized. To formalize the trade-off between these potential excesses we quantify the loss associated with a unit of
shortage by a constant $L>C$ (which assumes that the counter-measure has some practical value) and combine the total shortage loss with expenditure into a \emph{loss function}\footnote
{This does involve a confusing use of the word \emph{loss} to refer to two different quantities, but this seems to be
an ingrained and unavoidable habit in the literature.}
\[
l(x,y) = Cx + L(y-x)_{+}.
\] 
The decision problem is then to select an action $x$ that aligns with the preference that $l(x,y)$ be
as low as possible given any realization $Y=y$.  Equivalently, the problem is to choose a random future loss $l(x,Y)$ which
is \emph{admissible}, meaning that there is no alternative action $\tilde{x}$ producing a random future loss $l(\tilde{x},Y)$
which is never greater than $l(x,Y)$ and strictly lower for some realization $Y=y_0$.

Admissibility with respect to the loss $l$ is not, however, by itself a very strong decision criterion.
To give the decision problem more structure we assume the decision maker either knows the distribution $F$ of $Y$, or
wishes to proceed as if a forecast $F$ of $Y$ were true. This gives us what is known in decision theory as a decision
problem \emph{under risk} (regarding the future value of $Y$) as opposed to one \emph{under uncertainty} where both $Y$
as well as $F$ are unknown when the decision is to be made. A principle commonly invoked in this situation\footnote
{Note that this priciple might be inappropriate when the decision maker is \emph{risk averse} in some way such as
having a preference for random losses with lower variance.} is that the decision maker should or will seek to minimize the expected
loss
\begin{align}
  \mathbb{E}_{F}[l(x,Y)] = Cx + L\mathbb{E}_{F}[(Y-x)_{+}]. \label{eqn:expQloss}
\end{align}

@@

\section{Numerical computation of allocation Bayes acts}
\label{sec:numeric}

In practice, \eqref{eqn:quantiles-sum-to-K} is rarely amenable to closed form solution.
Using the Bayes act in explicit computations therefore requires a method of generating an approximation $\tilde{\tau}$ to 
a solution $\tau^{\star}$ of \eqref{eqn:tau-inclusion}.  One would also strive to have some control over a robust measure of 
of the approximation error of the random shortages which we aim to compute using the Bayes act. One possible measure is the sum
of absolute coordinate errors
\begin{align}
\mathrm{SAE}(Y,\tilde{\tau}) = \sum_{i=1}^{N} \left| (Y_i-q_{\tilde{\tau},F_i})_+ - (Y_i-x_i^{F,K})_+ \right|
\end{align}
% of the Bayes risk error or ``optimality gap''
% \begin{align}
% \left|\mathbb{E}_F\left[s_A(\bm{q}_{\tilde{\tau},F}, Y)\right] - \mathbb{E}_F[s_A(x^{F,K}, Y)]\right| = 
% L\left| \sum_{i=1}^{N}\mathbb{E}_{F_i}\left[(Y_i-q_{\tilde{\tau},F_i})_+ - (Y_i-x_i^{F,K})_+\right]\right|
% \end{align}
where $\bm{q}_{\tilde{\tau},F}$ solves the linear program
$\sum_{x \in \undertilde{\bm{F}}^{-1}(\tilde{\tau})}x_i = K$ approximating the true Bayes act defining linear program
$\sum_{x \in \undertilde{\bm{F}}^{-1}(\tau^{\star})}x_i = K$. In this section we outline such a method we have implemented 
in the \verb`R` package \verb`alloscore`.
